\Title

### 2023-07-21 Grimaldi-etal-BiolCybernetics.tex
sync

The precise timing of spikes emitted by neurons plays a crucial role in shaping the response of efferent biological neurons. This temporal dimension of neural activity holds significant importance in understanding information processing in neurobiology, especially concerning the output of neuromorphic hardware, such as event-based cameras. Nonetheless, many artificial neural models disregard this critical temporal dimension of neural activity. In this study, we present a model designed to efficiently detect temporal spiking motifs using a layer of spiking neurons equipped with heterogeneous synaptic delays. Our model capitalizes on the diverse synaptic delays present on the dendritic tree, enabling specific arrangements of temporally precise synaptic inputs to synchronize upon reaching the basal dendritic tree. We formalize this process as a time-invariant logistic regression, which can be trained using labeled data. To demonstrate its practical efficacy, we apply the model to naturalistic videos transformed into event streams, simulating the output of the biological retina or event-based cameras. To evaluate the robustness of the model in detecting visual motion, we conduct experiments by selectively pruning weights and demonstrate that the model remains efficient even under significantly reduced workloads. In conclusion, by providing a comprehensive, event-driven computational building block, the incorporation of heterogeneous delays has the potential to enhance the performance of future spiking neural network algorithms, particularly in the context of neuromorphic chips.

@@

In a recent review, we reported previous theoretical and  experimental evidence for the use of precise spiking motifs in biological neural networks~\cite{grimaldi_precise_2023}. In particular, \citet{abeles_role_1982} asked whether the role of cortical neurons is the integration of synaptic inputs or rather the detection of coincidences in temporal spiking motifs. While the first hypothesis favors the firing rate coding theory, the second emphasizes on the importance of temporal precision in neural activity. Since then, numerous studies have demonstrated the emergence of synchronous activity within a neuronal population~\citep{riehle_spike_1997, davis_spontaneous_2021}, efficient coding using precise spike timings~\citep{Perrinet2002,perrinet_coding_2004, gollisch_rapid_2008}, or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatiotemporal spiking motifs in biological spike trains. In neural models, the definition of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020} allows the efficient detection of these spatiotemporal motifs embedded in the spike train. Such spatiotemporal motifs present in neural activity may form useful representations to perform computations for various cognitive tasks using the synchrony of spikes reaching the soma of a neuron. In particular,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a spiking motif defined by a subset of neurons with different, but precise, relative spiking delays. This delay between a pair of connected neurons is defined as the time between the emission of a spike at the soma of the afferent neuron and its arrival at the soma of the efferent neuron. Importantly, due to the variety of weights and delays within a population, representations using polychronous groups have, in theory, a much higher information capacity than a firing rate-based approach.

@@

This paper aims to investigate the capability of the HD-SNN model to effectively learn and solve a motion detection task using realistic event-based data streams, as typically captured by event-based cameras, also known as Dynamic Vision Sensors (DVS). DVS are designed to mimic the signal transmitted from retinal ganglion cells to the visual cortex through the optic nerve. The events in these data streams are binary in nature and should provide sufficient information for performing fast and efficient motion detection. In this study, we first outline the task definition, specifying the requirements for motion detection. Subsequently, we describe the HD-SNN model employed for inferring motion, highlighting its key characteristics and architecture. Lastly, we elaborate on the training procedure employed to train the HD-SNN model for the specific motion detection task.

@@

To train and validate our model which uses supervised learning, we need to define a visual dataset for which we explicitly know the ground truth motion, that is, direction and speed. To achieve this, we define a procedure for animating a natural visual scene with virtual eye movements, similar to those used in  studies from neurobiology~\citep{vinje_sparse_2000, baudot_animation_2013} or computational neuroscience~\citep{kremkow_push-pull_2016}. First, we draw a trajectory inspired by biological eye movements. Indeed, these movements allow us to dynamically actuate the center of vision in the visual field. In animals with a fovea, this is particularly useful as it allows the area with the highest density of photoreceptors to be moved, for example, to a point of interest in the environment. A specific mechanism for this function are saccades, which are rapid eye movements that reposition the center of vision. In humans, saccades are very frequent (on average $2$ per second~\citep{dandekar_neural_2012}), very fast (about $80~\ms$), and have a wide range of speeds. On a more microscopic scale, the human gaze moves with minute micro-saccades which trajectories are similar to a Brownian trajectory~\citep{poletti_head-eye_2015}. To preserve the full generality of the task, we define eye movements using such a form of random walk~\citep{engbert_integrated_2011}: We first define a finite set of possible motions in polar coordinates as the Cartesian product of the $12$ regularly spaced directions and $3$ geometrically spaced speeds. Next, we define a saccadic path as a sequence of time segments whose durations are drawn from a Poisson distribution with a mean block length of $24~\ms$, similar to a Lévy flight~\citep[p. 289]{mandelbrot_fractal_1982}. Finally, assuming that  motion is stationary during each segment and that motions for each flight are drawn uniformly and independently, the global trajectory is generated by integrating this motion sequence. This generative model yields trajectories that are qualitatively very similar to those observed for human eye movements (\seeFig{motion_task}-\textit{(Left)}).

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}%
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \caption{
        {\bf Motion Detection Task.} To generate realistic event-based dynamic scenes, we mimic the effect of minute saccadic eye movements on a large natural scene ($1024\times1024$) by extracting an image ($128\times128$) which center is moving dynamically according to a jagged random walk. \textit{(Left)}~We show an instance of this trajectory (with a length of $200~\ms$, green line) superimposed on the luminance contrasts observed at time step $t=15~\ms$. \textit{(Right)}~The dynamics of this image, translated according to the saccadic trajectory, produces a naturalistic movie, which is then transformed into an event-based representation. We show snapshots of the resulting synthetic event stream at different time steps (from $t=15~\ms$ to $t=19~\ms$, these frames are marked on the trajectory by a white and black dot, respectively, in the left inset). Mimicking the response of ganglion cells in the retina, this representation encodes at each pixel all-or-none increases or decreases in luminance, i.e., ON (red) and OFF (blue) spikes. In the lower left corner of the snapshots, we show the corresponding instantaneous motion vector (red arrow). Note the change in the direction of motion between the third and fourth frames, and also that contours parallel to the motion produce fewer luminance changes, the so-called aperture problem, and thus relatively fewer spikes.
        }
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of $100$ large-scale natural images that were previously used to study the statistics of natural images~\citep{vanHateren1998}. Note that these were pre-processed to be in grayscale and to equalize (i.e., whiten) the energy in each frequency band, similarly to a process known to occur in the retino-thalamic pathway~\citep{dan_efficient_1996}. The full-scale images are $1024 \times 1024$ in size, and we crop images of size $128 \times 128$ positioned around the center of gaze at each time step. We discretize time in $1~\ms$ bins and produce movies of duration $N_T = 200~\ms$. To avoid boundary effects, we randomly position the full trajectory in image space so that the sub-image is translated using the position given by the trajectory at each time step and without touching the boundaries. At each time step, the translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the unit speed to correspond to a movement of one pixel per frame (i.e., per time bin of $1~\ms$).

@@

\subsection{The HD-SNN model}

@@

In biology, this new stream of events is naturally ordered in time as events reach the soma of postsynaptic neurons. In simulations, however, it should be properly reordered. Once transformed by the synaptic connectivity, this weighted event stream may be integrated, for instance as the membrane potential of a Leaky-Integrate-and-Fire neuron (\seeFig{izhikevich}-\textit{Left-Bottom}), yet the activation function of the HD-SNN neurons can be selected from the full range of spiking neuron response functions. Importantly, this activation function has to be such that when postsynaptic neurons are activated at their soma by a specific spatiotemporal motif imprinted in the synaptic set, and such that these spikes converge at the soma in a synchronous manner, the discharge probability should increase. In this subsection, we have briefly defined the HD-SNN model in all generality (see~\citep{Perrinet23ICANN} for a more specific description and treatment), and in the next subsection we describe an implementation of our model adapted to the motion detection task. 

@@

Such a method contrasts with classical methods for delay learning, which explicitly manipulate the delay as a variable and which are not directly differentiable~\citep{nadafian_bio-plausible_2020}. Keeping the analogy with spiking neurons, the analog activity $B$ represents the integration of synaptic activity, and we will now try to define the detection of motion using the spatiotemporal kernels. Since we know that at each instant, there may be different motions, we will define the activation function of our model as a sigmoid function that implements a form of Multinomial Logistic Regression (MLR). In our MLR model, a probability value for each class (i.e., each direction of motion) is predicted for each position $x, y$ and time $\timev$ as a sigmoid function $\sigma(\beta) = \frac {1} {1 + \exp(-\beta)}$ of the result of the convolution. Formally, using the kernels, the input raster plot is transformed into a probability with the following formula:
%
\begin{equation}\label{eq:mlr}
\forall x, y, \timev, \forall \class \in [1, \Nclass], \; \;
Pr(k=\class \; \vert \; x, y, \timev) =
\sigma (B(\class, \timev, x, y) +\bias_\class) 
\end{equation} 
%
where $\bias_\class$ is a scalar representing the bias associated with the class $\class$. In particular, we anticipate that certain specific patterns could result in closely synchronized outputs when they are integrated within the basal dendritic tree, consequently leading to heightened postsynaptic activity. By utilizing this analog representation of the evidence for each potential motion at every moment, we can progressively increase the likelihood of generating an output spike. To determine the spiking output, we establish a firing threshold. Here, we computed this threshold to ensure that neurons, on average, generate one spike per second. Therefore, the spiking output of the model corresponds to the motions in space and time that represent the highest probability.

Now that this general framework has been described formally, we may include some heuristics based on neuroscientific observations to constrain our model and its strategies for solving the ecological task described in section~\ref{sec:task}. Note that the general framework is an extension to that presented in~\cite{grimaldi_learning_2022}, in particular by including a more complex task, the deeper analysis of the results, and these novel neuroscience-inspired heuristics. First, to avoid introducing biases in the directions which may be learned, we apply a circular mask to the spatial dimensions of the kernels. We also included a prior in the selectable motions, as there is a prior for slow speeds in natural scenes~\citep{vacher_bayesian_2018}. Since we want to capture the possible convergence of the trajectories of the events converging on each voxel, we apply a mask to the spatiotemporal kernels such that the smaller the delay, the smaller the radius of the circular mask that is applied (\seeFig{kernels} for an illustration). In our simulations, we observed that including this prior accelerated the learning but was not necessary to reach convergence. %This is consistent with neuroscientific principles because, due to the limits of conduction delay, the synaptic delay is physically related to the distance between the presynaptic and postsynaptic neurons. 
Second, we observed that moving images produced trajectories of ON and OFF spikes, and that these were present in both polarities. This is due to the fact that our whitened images have a relative symmetry in the luminance profiles, that is, that an image with inverted contrast is indistinguishable from a standard one. Since this arrangement of polarities is independent of motion, we added a mechanism that collects the linear values for the movie and that with the ON and OFF cells flipped, keeping only the maximum value for each voxel. This is similar to the computation done for complex cells in primary visual cortex.

@@

If we now widen our focus on the interpretation of these kernels in terms of spatiotemporal motifs embedded in the event stream, these show a prototypical anisotropic profile adapted to motion detection~\citep{kaplan_anisotropic_2013}. In~\citep{grimaldi_robust_2022}, a link was drawn between event-based MLR training and Hebbian learning, allowing to say that the present model learns its weights according to a presynaptic activity associated to the different motion directions. Each neuron becomes selective to a specific motion direction through the learning of an associated prototypical spatiotemporal spiking motif. Each voxel in the 3D kernels defines a specific property by associating a weight to a position and a delay. Consequently, our model is able to detect precise spatiotemporal motifs embedded in the spike train and associated to the different motion directions. Note that as the delays become larger, two effects can be remarked. First, coefficients become lower which is consistent with the fact that trajectories are defined in a piece wise fashion, such that this decrease provides with an optimal integration considering the gradual diminishing of evidence as time progresses~\citep{Pasturel2020}. Second, coefficients become less localized compared to the kernel's spatial profile at short delays, consistent with the average diffusion of information included in the generative model and with the diffusion introduced in motion-based prediction models~\citep{perrinet_motion-based_2012,khoei_flash-lag_2017}. %This conic shape along the motion's trajectory highlights a loss of precision for longer delays, i.e. events away in the past.
\subsection{Accuracy versus efficiency trade-off}% for the motion detection task}
%: fig:accuracy
\begin{figure}%[h!]
    \includegraphics[width=0.95\linewidth]{figures/conv_HD-SNN.pdf}
        In response to a specific event-based input instance (Left), we present the neural activity of the HD-SNN model (Right). To aid visualization, we display a single spatio-temporal slice for a given vertical position ($y = 32$) and $10$ horizontal positions. The input spiking activity comprises ON and OFF spikes, as explained in Fig.~\ref{fig:motion_task}, showcasing the switching within the naturalistic event-based stream from one motion to another due to saccades. The dots above the graph indicate the corresponding motion class at each instant (the motion being represented by the matching color). The output activity consists of two components. Firstly, there is an analog component that corresponds to the evidence accumulated by the model on the spatio-temporal kernels. Secondly, there is a spiking component represented by vertical bars superimposed on the analog activity. These spikes signify moments when the evidence surpasses the spiking threshold. Importantly, this activity aligns with the motion depicted in the input stream. Finally, it is possible to compute the accuracy by comparing the ground truth motion in the input video with the motion predicted by the model (as represented by the colored dots on top of the graph).
    \label{fig:activity}
\end{figure}
After training our MLR model, we obtain spatiotemporal kernels corresponding to the weights associated with the heterogeneous delays of our layer of spiking neurons, which can be used for detection. For this, we quantify its ability to categorize different motions, i.e. on event streams for which the ground truth motion is known at each instant. When applied to new instances of the input movies, the model develops a neural activity which may be used to infer the correct motion~(\seeFig{activity}) and from which we may deduce an accuracy value. This accuracy was computed on a novel dataset of $200$ novel movies. The accuracy computed on the test set was approximately $91\%$ (with a chance level of $1 / 12/ 3 \approx 2\%$). %Moreover, we observed that the classification performance obtained for the model using time surfaces as in~\citep{grimaldi_robust_2022} gives similar results as our method.

\begin{figure}%[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/quant_accuracy.pdf}
    \caption{
        Accuracy as a function of computational load for the HD-SNN model (blue dots) with error bars indicating the 5\% - 95\% quantiles and a sigmoid fit (blue line). The relative computational load (on a logarithmic axis) is controlled by changing the percentage of nonzero weights relative to the dense convolution kernel. If we shorten the length of the kernel by using only the weights at the shortest delays, the accuracy quickly drops. However, if we prune the lowest coefficients from the whole kernel, we observe a stable accuracy value, with a drop to half-saturation observed at about $375$ times fewer computations.
        }
    \label{fig:accuracy}
\end{figure}
%
We also observed that the distribution of the kernel's weights is sparse, with most values close to zero (\seeFig{kernels}). As shown in the formalization of our event-based model, the computational cost of our model, if implemented on a neuromorphic chip, would be dominated by the computations used for the convolution operation. In a dense setting, this corresponds for all voxels in the output to a sum over all voxels in the inputs for all weights in the kernel. But if the information support is sparse, then computations can be now performed only on those events. Specifically, if we set some weights of the kernels to zero, then the additive operation in the convolution for those addresses can be dropped. As a consequence, computations will be performed only on those events which were multiplexed by the pruned connectivity matrix. Thus, knowing the sparseness of the input, the total number of computations scales with the number of spikes multiplied by the number of non-zero synaptic weights. This hypothesis is consistent with biological observations which have shown that communication consumes 35 times more energy than computation in the human cortex~\citep{Levy2021}.

In order to evaluate the resilience of the classification performance with respect to computational load, we adopt first a pruning approach, where we remove weights in $\kernel$ that fall below a specified threshold. The accuracy of classification is then plotted as a function of the relative number of computations or active weights per decision for each neuron in the layer (refer to Fig.~\ref{fig:accuracy}). To provide a basis for comparison and to account for the benefits of utilizing variable delays, we also present the accuracy achieved by an MLR model employing a shortening strategy. This strategy involves adjusting the temporal width by selecting only the weights associated with the shortest delays. In comparison to the inference performed using the complete 3D kernels without any pruning ($36\times2\times21\times17\times17$), both approaches demonstrate a reduction in computational requirements as indicated by the number of non-zero weights.

>>>>

The precise timing of spikes from afferent neurons determines the response of a biological neuron. This temporal aspect of neural activity is essential for understanding information processing in neurobiology, and is particularly applicable to the output of neuromorphic hardware such as event-based cameras. However, most artificial neural models ignore this important temporal dimension of neural activity. Here, we develop a model for efficiently detecting temporal spiking motifs based on a layer of spiking neurons with heterogeneous synaptic delays. The model exploits the property that the diversity of synaptic delays on the dendritic tree allows specific arrangements of temporally precise synaptic inputs to synchronize as they reach the basal dendritic tree. We show that this process can be formalized as a time-invariant logistic regression that can be trained on labeled data. We demonstrate its application to naturalistic videos transformed into event streams, similar to the output of the biological retina or event-based cameras. To characterize the robustness of the model in detecting visual motion, we test its efficiency by pruning some weights and show that it is still efficient at very low workloads. In conclusion, by providing an end-to-end, event-driven computational building block, the learning of heterogeneous delays could improve the performance of future spiking neural network algorithms, and in particular their applications in neuromorphic chips.

@@

In a recent review, we reported previous theoretical and  experimental evidence for the use of precise spiking motifs in biological neural networks~\cite{grimaldi_precise_2023}. In particular, \citet{abeles_role_1982} asked whether the role of cortical neurons is the integration of synaptic inputs or rather the detection of coincidences in temporal spiking motifs. While the first hypothesis favors the firing rate coding theory, the second emphasizes on the importance of temporal precision in neural activity. Since then, numerous studies have demonstrated the emergence of synchronous activity within a neuronal population~\citep{riehle_spike_1997, davis_spontaneous_2021}, efficient coding using precise spike timings~\citep{Perrinet2002,perrinet_coding_2004, gollisch_rapid_2008}, or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatiotemporal spiking motifs in biological spike trains. In neural models, the definition of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020} allows the efficient detection of these spatiotemporal motifs embedded in the spike train. These spatiotemporal motifs present in neural activity may form useful representations to perform computations for various cognitive tasks using the synchrony of spikes reaching the soma of a neuron. In particular,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a spiking motif defined by a subset of neurons with different, but precise, relative spiking delays. This delay between a pair of connected neurons is defined as the time between the emission of a spike at the soma of the afferent neuron and its arrival at the soma of the efferent neuron. Importantly, due to the variety of weights and delays within a population, representations using polychronous groups have, in theory, a much higher information capacity than a firing rate-based approach.

@@

In this paper, we test whether the HD-SNN model is able to efficiently learn to solve a motion detection task defined by a realistic event-based data stream. This type of signal is commonly captured by an event-based camera, or Dynamic Vision Sensor (DVS), and is inspired by the signal sent from the retinal ganglion cells through the optic nerve to the visual cortex. The events are binary in nature and should be sufficient to perform the motion detection task in a fast and efficient manner. Here, we first describe the task definition, then the HD-SNN model we use for motion inference, and finally how we train it.

@@

To train and validate our model which uses supervised learning, we need to define a visual dataset for which we explicitly know the ground truth motion, that is, direction and speed. To achieve this, we define a procedure for animating a natural visual scene with virtual eye movements, similar to those used in  studies from neurobiology~\citep{vinje_sparse_2000, baudot_animation_2013} or computational neuroscience~\citep{kremkow_push-pull_2016}. First, we draw a trajectory inspired by biological eye movements. Indeed, these movements allow us to dynamically actuate the center of vision in the visual field. In animals with a fovea, this is particularly useful as it allows the area with the highest density of photoreceptors to be moved, for example, to a point of interest in the environment. A specific mechanism for this function are saccades, which are rapid eye movements that reposition the center of vision. In humans, saccades are very frequent (on average $2$ per second~\citep{dandekar_neural_2012}), very fast (about $80~\ms$), and have a wide range of speeds. On a more microscopic scale, the human gaze moves with minute micro-saccades which trajectories are similar a Brownian trajectory~\citep{poletti_head-eye_2015}. To preserve the full generality of the task, we define eye movements using such a form of random walk~\citep{engbert_integrated_2011}: We first define a finite set of possible motions in polar coordinates as the Cartesian product of the $12$ regularly spaced directions and $3$ geometrically spaced speeds. Next, we define a saccadic path as a sequence of time segments whose durations are drawn from a Poisson distribution with a mean block length of $24~\ms$, similar to a Lévy flight~\citep[p. 289]{mandelbrot_fractal_1982}. Finally, assuming that  motion is stationary during each segment and that motions for each flight are drawn uniformly and independently, the global trajectory is generated by integrating this motion sequence. This generative model yields trajectories that are qualitatively very similar to those observed for human eye movements (\seeFig{motion_task}-\textit{(Left)}).

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}%
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \caption{
        {\bf Motion Detection Task.} To generate realistic event-based dynamic scenes, we mimic the effect of minute saccadic eye movements on a large natural scene ($1024\times1024$) by extracting an image ($128\times128$) which center is moving dynamically according to a jagged random walk. \textit{(Left)}~We show an instance of this trajectory (with a length of $200~\ms$, green line) superimposed on the luminance contrasts observed at time step $t=15~\ms$. \textit{(Right)}~The dynamics of this image, translated according to the saccadic trajectory, produces a naturalistic movie, which is then transformed into an event-based representation. We show snapshots of the resulting synthetic event stream at different time steps (from $t=15~\ms$ to $t=19~\ms$, these frames are marked on the trajectory by a white and black dot, respectively, in the left inset). Mimicking the response of ganglion cells in the retina, this representation encodes at each pixel all-or-none increases or decreases in luminance, i.e., ON (red) and OFF (blue) spikes. In the lower left corner of the snapshots, we show the corresponding instantaneous motion vector (red arrow). Note the change in the direction of motion between the third and fourth frames, and also that contours parallel to the motion produce fewer luminance changes, the so-called aperture problem, and thus relatively fewer spikes.
        }
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of $100$ large-scale natural images that were previously used to study the statistics of natural images~\citep{vanHateren1998}. Note that these were pre-processed to be in grayscale and to equalize (i.e., whiten) the energy in each frequency band, similarly to a process known to occur in the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $1024 \times 1024$ in size, and we crop images of size $128 \times 128$ positioned around the center of gaze at each time step. We discretize time in $1~\ms$ bins and produce movies of duration $N_T = 200~\ms$. To avoid boundary effects, we randomly position the full trajectory in image space so that the sub-image is translated using the position given by the trajectory at each time step and without touching the boundaries. At each time step, the translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the unit speed to correspond to a movement of one pixel per frame (i.e., per time bin of $1~\ms$).

@@

\subsection{The Heterogeneous Delays SNN (HD-SNN) model}

@@

In biology, this new stream of events is naturally ordered in time as events reach the soma of postsynaptic neurons. In simulations, however, it should be properly reordered. Once transformed by the synaptic connectivity, this weighted event stream may be integrated, for instance as the membrane potential of a Leaky-Integrate-and-Fire neuron (\seeFig{izhikevich}-\textit{Left-Bottom}), yet the activation function of the HD-SNN neurons can be selected from the full range of spiking neuron response functions. Importantly, this activation function has to be such that when postsynaptic neurons are activated at their soma by a specific spatiotemporal motif imprinted in the synaptic set, and such that these spikes converge at the soma in a synchronous manner, the discharge probability should increase. In this subsection, we have briefly defined the HD-SNN model in all generality (see~\citep{KeatingPerrinet23ICANN} for a more specific description and treatment), and in the next subsection we describe an implementation of our model adapted to the motion detection task. 

@@

Such a method contrasts with classical methods for delay learning, which explicitly manipulate the delay as a variable and which are not directly differentiable~\citep{nadafian_bio-plausible_2020}. Keeping the analogy with spiking neurons, the analog activity $B$ represents the integration of synaptic activity, and we will now try to define the detection of motion using the spatiotemporal kernels. Since we know that at each instant, there may be different motions, we will define the activation function of our model as a sigmoid function that implements a form of Multinomial Logistic Regression (MLR). In our MLR model, a probability value for each class (i.e., each direction of motion) is predicted for each position $x, y$ and time $\timev$ as a sigmoid function $\sigma(\beta) = \frac {1} {1 + \exp(-\beta)}$ of the result of the convolution. Formally, using the kernels, the input raster plot is transformed into a probability with the following formula:
%
\begin{equation}\label{eq:mlr}
\forall x, y, \timev, \forall \class \in [1, \Nclass], \; \;
Pr(k=\class \; \vert \; x, y, \timev) =
\sigma (B(\class, \timev, x, y) +\bias_\class) 
\end{equation} 
%
where $\bias_\class$ is a scalar representing the bias associated with the class $\class$. In particular, we anticipate that certain specific patterns could result in closely synchronized outputs when they are integrated within the basal dendritic tree, consequently leading to heightened postsynaptic activity. By utilizing this analog representation of the evidence for each potential motion at every moment, we can progressively increase the likelihood of generating an output spike. To determine the spiking output, we establish a firing threshold. We computed this threshold to ensure that neurons, on average, generate one spike per second. Therefore, the spiking output of the model corresponds to the motions in space and time that represent the highest probability.

Now that this general framework has been described formally, we may include some heuristics based on neuroscientific observations to constrain our model and its strategies for solving the ecological task described in section~\ref{sec:task}. Note that the general framework is an extension to that presented in~\cite{grimaldi_learning_2022}, in particular by including a more complex task, the deeper analysis of the results, and these novel neuroscience-inspired heuristics. First, to avoid introducing biases in the directions which may be learned, we apply a circular mask to the spatial dimensions of the kernels. We also included a prior in the selectable motions, as there is a prior for slow speeds in natural scenes~\citep{vacher_bayesian_2018}. Since we want to capture the possible convergence of the trajectories of the events converging on each voxel, we apply a mask to the spatiotemporal kernels such that the smaller the delay, the smaller the radius of the circular mask that is applied (\seeFig{kernels} for an illustration). %This is consistent with neuroscientific principles because, due to the limits of conduction delay, the synaptic delay is physically related to the distance between the presynaptic and postsynaptic neurons. 
Second, we observed that moving images produced trajectories of ON and OFF spikes, and that these were present in both polarities. This is due to the fact that our whitened images have a relative symmetry in the luminance profiles, that is, that an image with inverted contrast is indistinguishable from a standard one. Since this arrangement of polarities is independent of motion, we added a mechanism that collects the linear values for the movie and that with the ON and OFF cells flipped, keeping only the maximum value for each voxel. This is similar to the computation done for complex cells in primary visual cortex.

@@

If we now widen our focus on the interpretation of these kernels in terms of spatiotemporal motifs embedded in the event stream, these show a prototypical anisotropic profile adapted to motion detection~\citep{kaplan_anisotropic_2013}. In~\citep{grimaldi_robust_2022}, a link was drawn between event-based MLR training and Hebbian learning, allowing to say that the present model learns its weights according to a presynaptic activity associated to the different motion directions. Each neuron becomes selective to a specific motion direction through the learning of an associated prototypical spatiotemporal spiking motif. Each voxel in the 3D kernels defines a specific property by associating a weight to a position and a delay. Consequently, our model is able to detect precise spatiotemporal motifs embedded in the spike train and associated to the different motion directions. Note that as the delays become larger, two effects can be remarked. First, coefficients become lower which is consistent with the fact that trajectories are defined in a piece wise fashion, such that this decrease provides with an optimal integration considering the gradual diminishing of evidence as time progresses~\citep{Pasturel2020}. Second, coefficients become less localized compared to the kernel's spatial profile at short delays, consistent with the average diffusion of information included in the generative model and with the diffusion introduced in motion-based prediction models~\citep{perrinet_motion-based_2012,khoei_flash-lag_2017}. %This conic shape along the motion's trajectory highlights a loss of precision for longer delays, i.e. events away in the past.
%
%
\subsection{Accuracy versus efficiency trade-off}% for the motion detection task}
%%%-----------------------------------------------------------------
%
%%%-----------------------------------------------------------------
%: fig:accuracy
%%%-----------------------------------------------------------------
\begin{figure}%[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/quant_accuracy.pdf}
    \caption{
        Accuracy as a function of computational load for the HD-SNN model (blue dots) with error bars indicating the 5\% - 95\% quantiles and a sigmoid fit (blue line). The relative computational load (on a logarithmic axis) is controlled by changing the percentage of nonzero weights relative to the dense convolution kernel. As we prune the coefficients, we observe a stable accuracy value, with a drop observed at about $250$ times fewer computations.
        }
    \label{fig:accuracy}
\end{figure}
%
After training our MLR model, we obtain spatiotemporal kernels corresponding to the weights associated with the heterogeneous delays of our layer of spiking neurons, which can be used for detection. For this, we quantify its ability to categorize different motions, i.e. on event streams for which the ground truth motion is known at each instant. The accuracy was computed on a novel dataset of $200$ novel movies. The accuracy computed on the test set was approximately $91\%$ (with a chance level of $1 / 12/ 3 \approx 2\%$). %Moreover, we observed that the classification performance obtained for the model using time surfaces as in~\citep{grimaldi_robust_2022} gives similar results as our method.

We also observed that the distribution of the kernel's weights is sparse, with most values close to zero (\seeFig{kernels}). As shown in the formalization of our event-based model, the computational cost of our model, if implemented on a neuromorphic chip, would be dominated by the computations used for the convolution operation. In a dense setting, this corresponds for all voxels in the output to a sum over all voxels in the inputs for all weights in the kernel. But if the information support is sparse, then computations can be now performed only on those events. Specifically, if we set some weights of the kernels to zero, then the additive operation in the convolution for those addresses can be dropped. As a consequence, computations will be performed only on those events which were multiplexed by the pruned connectivity matrix. Thus, knowing the sparseness of the input, the total number of computations scales with the number of spikes multiplied by the number of non-zero synaptic weights. This hypothesis is consistent with biological observations which have shown that communication consumes 35 times more energy than computation in the human cortex~\citep{Levy2021}.

In order to evaluate the resilience of the classification performance with respect to computational load, we adopt first a pruning approach, where we remove weights in $\kernel$ that fall below a specified threshold. The accuracy of classification is then plotted as a function of the relative number of computations or active weights per decision for each neuron in the layer (refer to Fig.~\ref{fig:accuracy}). To provide a basis for comparison and to account for the benefits of utilizing variable delays, we also present the accuracy achieved by an MLR model employing a shortening strategy. This strategy involves adjusting the temporal width by selecting only the weights associated with the shortest delays. In comparison to the inference performed using the complete 3D kernels without any pruning ($21\times17\times17$), both approaches demonstrate a reduction in computational requirements as indicated by the number of non-zero weights.

By selectively setting certain weights to zero, we observe that the accuracy's evolution, as a function of the logarithmic percentage of active weights, aligns well with a sigmoid curve for both pruning and shortening strategies. The shortening strategy (depicted in orange) demonstrates a rapid decline in accuracy, reaching half-saturation when approximately one-third of the weights are removed. On the other hand, the pruning strategy (shown in blue) exhibits a different behavior. It reaches half-saturation when the ratio of active weights is approximately $2.6\times 10^{-3}$, corresponding to around $375$ less computations compared to the dense scenario. In comparison to using the complete kernels, our method maintains accuracy close to its peak performance even when the number of computations is divided by a factor of up to around 31. This substantial reduction in computations showcases the robustness of the presented method.
%
%
%%%-----------------------------------------------------------------
%
\subsection{Testing with natural-like textures}
%%%-----------------------------------------------------------------
%%%-----------------------------------------------------------------
%: fig:motion_clouds
%%%-----------------------------------------------------------------
\begin{figure}%[h!]
    \centering
    \includegraphics[width=0.99\linewidth]{figures/motion_clouds.pdf} %
    % TODO : show prototypes of textures on top
    \caption{{\bf Role of stimulus parameters in motion detection accuracy.} Accuracy as a function of (from left to right) mean spatial frequency, bandwidth in spatial frequency (from a grating (left) to isotropic textures (right)), bandwidth in orientation (from isotropic textures (left) to grating (right)), bandwidth in speed (from a rigid motion (left) to independent frames (right)). Examples snapshots are shown as an illustration in the top insets. Note that these accuracies are computed both in the case where the orientation of the synthetic texture is necessarily perpendicular to the motion ('no aperture' condition) and in the generic case where the orientation is independent of direction ('aperture').
    }
    \label{fig:motion_clouds}
\end{figure}
%%%-----------------------------------------------------------------
In order to assess the influence of spatiotemporal parameters of the stimuli on the performance of the model, we now test the model on simpler, parameterized stimuli. For this purpose, we use a set of synthetic visual stimuli, \textit{Motion Clouds}~\citep{leon_motion_2012}, which are natural-like random textures for which we can control relevant parameters for motion detection, including motion direction, spatial orientation, and spatial frequency along with their respective precisions (\seeFig{motion_clouds})~\citep{Leon2012,vacher_bayesian_2018}. By matching the spatial and temporal characteristics of the generated movies with those of the motion task mentioned earlier, we created a range of textures featuring different spatial properties and motions. This procedure defines a set of textures with different spatial properties and different motions chosen from the same set of $12$ directions and $3$ speeds. For each motion, we also varied the texture parameters, such as mean and variance of orientation or spatial frequency content, to provide some naturalistic variability. This method provides a rich dataset of textured movies for which we know the ground truth for the motion.
